Double Standards in Judging Collective Action

### 2022-09-13 Documents/manuscript.Rmd
preprint

Protest movements often spark contention about what actions are acceptable means of protest. Reactions to Black American athletes kneeling during the national anthem to protest racist police violence are a case in point. Kneeling during the national anthem is not violent, disruptive, or illegal. And yet, only 29\% of White Americans (66\% of Black Americans) and 11\% of Republicans (59\% of Democrats) considered it appropriate for Black athletes to protest in this way [@yougov_nfl_2017]. Indeed, Black athletes who had participated in the protests faced a backlash in public opinion [@lacina_nfl_2020]. 
<!-- Likewise, 42\% of Americans thought that, in the 2020 Black Lives Matter protests, most protesters were trying to incite violence or destroy property, even though very few protests involved any violence or property destruction [@acled_year_2021]. -->
This and other examples suggest that *who* the protesters are and *what* they are protesting influences how acceptable observers judge their actions to be.

@@

<!-- In the next sections, we, first, introduce the consequential distinction between normative and nonnormative collective action and define double standards in judging what actions are acceptable and unacceptable means to effect social change. We, second, derive predictions from social psychological theories about identity- and ideology-based double standards that might affect judgements about collective action. We, third, introduce item response theory as a conceptual and statistical framework for study double standards in judging collective action. We, finally, provide an overview of the present research. -->

@@

Collective action can take many forms. Psychologists distinguish between *normative* collective action that seeks to achieve a political goal while conforming to the norms of the existing social system and *non-normative* collective action that violates those norms [@wright_responding_1990]. Recent research has shown that this distinction matters for how people respond to collective action. @shuman_disrupting_2021 showed that non-normative, non-violent collective action is more effective than either normative or violent collective action at gaining concessions from those most resistant to social change. @feinberg_activists_2020 demonstrated, however, that observers are less supportive of social movements that use extreme protest actions. Similarly, @teixeira_is_2020 found that advantaged-group members perceive non-normative collective action by disadvantaged-group members as more damaging to their ingroup's social image and are thus less supportive of such action. @verkuyten_intolerance_2022 found that, in general, the public tends to be intolerant of non-normative collective action. This means that activists face the dilemma that non-normative collective action might be most effective at gaining concessions but that it also risks reducing popular support for a cause. Together, these studies show that the distinction between normative and non-normative collective action is psychologically and societally consequential.

Research has, for the most part, relied on ad-hoc distinctions between what researchers themselves considered normative and non-normative collective action. In liberal democracies, researchers tend to consider actions such as signing petitions, voting in elections, or peaceful protest to be normative and actions such as blocking traffic, damaging property, or violent protest to be non-normative. But, as the examples in the first paragraph show, the distinction between normative and non-normative collective action might instead be in the eye of the beholder.

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable and unacceptable forms of collective action. For that purpose, we define a double standard as judging the same action as more or less acceptable depending on who is performing the action and for what reason [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.] In this section, we derive predictions from social psychological theories about how identity-based and ideology-based double standards could lead people to judge the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting.

@@

One domain in which ingroup bias manifests is judgements about other people's actions. @hewstone_ultimate_1990 reviewed research showing that people make ingroup-serving causal attributions when judging actions by ingroup and outgroup members, for example, by attributing negative behavior by outgroup members to internal causes but attributing negative behavior by ingroup members to external causes. @valdesolo_moral_2007 demonstrated that participants judged the same selfish action as less unfair when performed by themselves or an ingroup member than when performed by an outgroup member. Other studies [@abrams_double_2013; @endevelt_everyone_2021; for exceptions, see @mendoza_for_2014; @pinto_membership_2010] provided further evidence for ingroup bias in judging moral transgressions by ingroup and outgroup members.

@@

System Justification Theory [@jost_role_1994; for a recent review, see @jost_theory_2020] argues that people are motivated to defend, justify, and bolster the prevailing social, economic, and political system because doing so serves basic epistemic, existential, and relational needs. Both advantaged and disadvantaged groups are thought to be motivated to justify the existing system, although the strength of this motivation and its expression vary across individuals and situations.

@@

Just as system justification shapes support for, and opposition to, collective action, we propose that it results in ideology-based double standards in what actions people consider acceptable forms of collective action. First, we hypothesize that  people will judge the same protest actions as more acceptable if the protest supports, rather than challenges, the system (*Hypothesis 2a*). This hypothesis reflects the assumption that the motivation to justify the system is, to some extent, universal. Second, we hypothesize that people will judge the same protest actions as more acceptable if they endorse system-justifying beliefs and the protest supports the system or if they reject system-justifying beliefs and the protest challenges the system (*Hypothesis 2b*). Notably, this hypothesis implies a symmetry in how people with different ideological orientations judge system-supporting and system-challenging collective action.

@@

Item response theory is a a conceptual and statistical framework for understanding how the characteristics of both items and respondents shape responses to a set of items [@demars_irt_2010]. 
<!-- As such, it allows us to disentangle how acceptable protest actions are from how accepting respondents are of them and, in so doing, to detect double standards in judging collective action.  -->
When respondents judge whether each of a set of actions is an acceptable means of protest (1 = *yes*, 0 = *no*), a two parameter logistic item response theory model estimates responses as a function of several latent (unobserved) parameters: 
$$ \Pr ( y_{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) $$
where $\Pr ( y_{ij} = 1 )$ is the probability that respondent $j$ considers action $i$ an acceptable means of protest, $\theta_j$ estimates how *accepting* respondent $j$ is of various protest actions, $\beta_i$ estimates how *acceptable* action $i$ is, and $\alpha_i$ estimates how *discriminating* action $i$ is.

@@

In this research, item response theory serves two purposes. First, item response theory provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in the experiments. Second, item response theory provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters and what they are protesting. In Experiments 1--3, we operationalize double standards as the differences in $\theta_{(z)}$ between conditions which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes.

@@

The first purpose of our research is to develop an instrument to measure where people draw the line between acceptable and unacceptable forms of collective action. In Study 1, we ask participants to generate more and less extreme protest actions and compile a pool of protest actions for scale development. In Study 2, we ask participants to rate how acceptable they consider each of those protest action to be and use item response theory to select the most discriminating and informative protest actions for our instrument. In so doing, we build an instrument that enables both present and future research to capture double standards in judging collective action. 
<!-- In this way, our research situates the distinction between normative and non-normative collective action in the eye of the beholders and opens it to scientific investigation. -->

@@

<!-- > **Hypothesis 1.** People will consider *more* controversial actions acceptable means of protest if (a) the protesters are ingroup members *or* if (b) the cause of the protest aligns with the ingroup's interests. -->

<!-- > **Hypothesis 2.** People will consider *more* controversial actions acceptable means of protest if (a) the protest's cause supports, rather than challenges, the system *or* if (b) the protest's cause aligns with their system-justifying beliefs. -->

@@

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that, in line with *Hypothesis 1a* and *1b*, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members. Assuming that collective action by lower-status group members is seen as more threatening to the unequal status quo than collective action by higher-status group members and that people are, to same extent, motivated to justify the system, we tested the preregistered hypothesis that, in line with *Hypothesis 2a*, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group than if they were from a lower-status group. In non-preregistered analyses, we further explored whether, in line with *Hypothesis 2b*, participants would judge the same protest actions as more or less acceptable depending on their own ideological positions. In this way, Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

We recruited 515 participants from the Prolific subject pool who were UK citizens, 25 years old or older, and not current students.^[We had preregistered a sample of 500 participants, before exclusions, but included another 15 participants who completed the study without returning an approval code to the Prolific platform. Participants received, on average, \pounds 9.00 (\$10.93) per hour of participation.] As preregistered, we excluded 71 participants who failed an attention check. This resulted in a final sample of 443 participants ($\textit{Mdn} = 41$ years, age range: 25--76 years; 272 women, 171 men, 1 non-binary). Of these, 210 participants considered their past, current, or future jobs to be working-class jobs. Participants in this lower-status group did not have a university degree and placed themselves on the bottom three ranks of the subjective socioeconomic status ladder. Another 233 participants considered their past, current job, or future jobs to be middle-class/professional jobs. Participants in this higher-status group had at least an undergraduate degree and placed themselves on the top four ranks of the subjective socioeconomic status ladder.

@@

We assessed reactions to the vignette by asking participants how outraged they would be if the government were to introduce this bill in real life, to what extent it would affect them personally, and to what extent the it would affect people like them (1 = *not at all*, 7 = *very much*). We also asked participants to what extent they identified with people with the kinds of jobs most affected by the proposed bill (1 = *not at all*, 7 = *very much*) and how often, if at all, they had participated in protest actions such as the ones we asked about (1 = *never*, 5 = *very often*).

@@

Figure \ref{fig:f3} (B) the $z$-standardized propensity to consider more controversial protest actions acceptable as a function of the three ideological orientation variables. We found that participants who reported a more right-wing political orientation ($\beta_{xy} = -0.25, [-0.37, -0.15]$) and who expressed more agreement with system-justifying beliefs ($\beta_{xy} = -0.20, [-0.30, -0.10]$) tended to find fewer collective actions to be acceptable. In contrast, we found that, after controlling for the other two variables, social dominance orientation was not associated with participants' judgements about how acceptable various collective actions are ($\beta_{xy} = -0.05, [-0.15, 0.05]$). Overall, these findings suggest that people who are right-wing and endorse system-justifying beliefs tend to find various collective actions to be less acceptable than people who are left-wing and do not endorse system-justifying beliefs.^[We ran additional analyses to explore how these associations differed across experimental conditions. We found uncertainty intervals for these associations to overlap across conditions, suggesting that we did not have enough data to differentiate these varying effects.]

@@

Experiment 1 did not, however, provide a conclusive test of our hypotheses. First, by focusing on fictitious government bills, we might have chosen a scenario too far removed from current affairs to evoke strong ingroup bias in participants' judgements of protest actions. That said, participants' reactions to the experimental manipulation suggested that it evoked outrage and was understood in ingroup--outgroup terms. Second, by focusing on collective action to defend workers' rights, we arguably studied reactions to collective action for a progressive cause. To test for ideology-based double standards, we relied on the assumption that collective action by lower-status group members would be seen as more challenging to the prevailing system than collective action by higher-status group members. This assumption, however, was contradicted by our observation that participants who endorsed system-justifying beliefs tended to find actions by lower-status and higher-status protesters to be less acceptable. Experiment 2 addressed those limitations by focusing on a scenario we expected to evoke stronger responses from participants and by examining reactions to both system-challenging and system-defending collective action.

@@

In Experiment 2, we investigated potential double standards in judging collective action for or against defunding the police in the United States. That is, we tested whether Black and White Americans applied different standards when judging collective action taken by either Black or White protesters to protest either for or against police divestment as a possible solution to racist police violence. We conducted this study in January 2021, a time when, after a year of unprecedented protests for racial justice, most Americans could be expected to be aware of, and have formed an opinion on, the Black Lives Matter movement. We focused on defunding the police as this position remained controversial among both Black and White Americans even as they broadly agreed on the need for police reform.^[In December 2020, only 34% of Black Americans and 18% of White Americans supported defunding the police while majorities of Black and White Americans supported police reforms such as eliminating qualified immunity (50% and 57%) or banning chokeholds [79% and 61%, @yougov_police_2020].]

@@

As expected, Black participants ($M = 6.18, \textit{SD} = 1.31$) reported being more outraged about recent incidents of police violence than White participants ($M = 5.56, \textit{SD} = 1.61$; Cohen's $d = 0.42$). Black participants strongly identified with their ingroup ($M = 6.46, \textit{SD} = 1.09$) and, across experimental conditions, identified more with Black protesters ($M = 4.66, \textit{SD} = 1.99$) than with White protesters ($M = 4.03, \textit{SD} = 2.16$; $d = 0.30$). In contrast, White participants identified less strongly with their ingroup ($M = 4.85, \textit{SD} = 1.53$; $d = 1.21$) and did not identify more with White protesters ($M = 3.58, \textit{SD} = 2.01$) than with Black protesters ($M = 3.81, \textit{SD} = 1.77$; $d = -0.12$). Both White ($d = 0.67$) and Black ($d = 0.53$) participants tended to identify more with protesters protesting *for* defunding the police. On average, participants tended to describe their political orientation as moderately liberal ($M = 2.92, \textit{SD} = 1.73$) but, as expected, were divided in their support for defunding the police ($M = 3.37, \textit{SD} = 1.42$). That is, 56% supported defunding the police, 30% opposed defunding the police, and 14% remained undecided.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a function of their *z*-standardized endorsement of system-justifying beliefs. As preregistered, we modeled this relationship with two fixed effects, one estimating the effect of system-justifying beliefs on judgements about system-defending protest actions and one estimating their effect on judgements about system-challenging protest actions, and one varying effect estimating its variance across conditions. Supporting *Hypothesis 2b*, participants who *rejected* system-justifying beliefs were *more* likely to consider system-challenging protest actions (for defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "for"])`). We did not, however, find evidence for the ideological symmetry implied by *Hypothesis 2b* since participants who *endorsed* system-justifying beliefs were *not* more likely to consider system-defending protest actions (against defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "against"])`).

@@

We explored whether group identification moderated how the participants' and the protesters' group memberships affected participants' responses. To that end, we extended Model 1 by estimating participants' responses as a function of their *z*-standardized identification with their racial ingroup. As in Model 2, we modeled this relationship with a fixed and a varying effect. We found, however, that even participants who strongly identified with their racial ingroup (+1*SD*) did not consider protest actions performed by their ingroup to be more acceptable than the same actions performed by the relevant outgroup (`r with(m5, d_text[contrast == "h1a_contrast" & x_ide == 1L])`) and did not consider protest actions for a cause that was aligned with their ingroup's interests to be more acceptable than protest actions for a cause not aligned with their ingroup's interests (`r with(m5, d_text[contrast == "h1b_contrast" & x_ide == 1L])`). Our non-preregistered analyses thus suggested that group identification did not moderate group differences in judgements about collective action by ingroup and outgroup members.

@@

In Experiment 3, we investigated potential double standards in judging collective action for or against restricting abortion in the United States. That is, we tested whether liberal and conservative women and men applied different standards when judging collective action opposing or supporting further restricting or banning abortion. We conducted this study in June 2022 when many expected the Supreme Court to overturn its *Roe v. Wade* decision that hitherto had guaranteed access to legal abortions in the first 20--24 weeks of pregnancy in the United States. We focused on restricting abortion as it is a conservative issue, as it primarily affects women's reproductive rights, and as the anticipated Supreme Court decision would prompt several states to restrict or ban abortion.

By varying the participants' political orientation and the protesters' cause, we tested the preregistered hypothesis that, in line with *symmetrical* ideology-based double standards, conservatives would judge the same protest actions to be more acceptable if the protesters supported restricting abortion and that liberals would judge the same protest actions to be more acceptable if the protesters opposed restricting abortion. We tested the alternative hypothesis that, in line with *asymmetrical* ideology-based double standards, conservatives would judge all protest actions to be equally unacceptable while liberals would judge the same protest actions to be more acceptable if the protesters opposed restricting abortion. By varying the participants' gender, we further tested the preregistered hypothesis that, in line with identity-based double standards, female participants would judge protest actions against restricting abortion to be more acceptable, on average, than male participants or protest actions for restricting abortion.

@@

We preregistered the sample size, hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822](https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822)). We deviated from the preregistration in two ways. First, we stopped collecting data before reaching the preregistered sample size. We did so as the Supreme Court decision on reproductive rights, announced on June 24, 2022, had rendered the experimental manipulation invalid. Second, we do not report results for two further hypotheses---testing moral conviction as a potential mechanism underlying ideology-based double standards---as those findings are beyond the scope of this manuscript and will be reported in a future publication. We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)).

@@

We recruited 804 participants from the Prolific subject pool who were 18 years old or older and lived in the United States.^[Data were collected between June 10 and 23, 2022. Participants received, on average, \$24.70 per hour of participation.] We balanced the sample's gender and political composition so that half of all participants identified as conservative and the other half identified as liberal and so that roughly half of the participants in each group were women and men. As preregistered, we excluded 71 participants who failed at least one of five attention checks or reported a different political orientation than they had reported in the Prolific prescreening questionnaire. As noted above, we had to stop collecting data before reaching the preregistered sample size of 800 eligible participants. This resulted in a final sample of 733 participants ($\textit{Mdn} = 37$ years, age range: 18--93 years) of whom 181 were conservative men, 190 were conservative women, 185 were liberal men, and 177 were liberal women.

@@

After answering where they would you place themselves along the political spectrum, in line with their response in the Prolific prescreening questionnaire, participants read the following paragraph:

>>>>

Protest movements often spark contention about what actions are acceptable means of protest. Reactions to Black American athletes kneeling during the national anthem to protest racist police violence are a case in point. Kneeling during the national anthem is not violent, disruptive, or illegal. And yet, only 29\% of White Americans (compared to 66\% of Black Americans) and 11\% of Republicans (compared to 59\% of Democrats) considered it appropriate for Black athletes to protest in this way [@yougov_nfl_2017]. Indeed, Black athletes who had participated in the protests faced a backlash in public opinion [@lacina_nfl_2020].<!-- Likewise, 42\% of Americans thought that, in the 2020 Black Lives Matter protests, most protesters were trying to incite violence or destroy property, even though very few protests involved any violence or property destruction [@acled_year_2021]. --> This and other examples suggest that *who* the protesters are and *what* they are protesting influences how acceptable observers judge their actions to be.

@@

@@

Collective action can take many forms. Psychologists distinguish between *normative* collective action that seeks to achieve a political goal while conforming to the norms of the existing social system and *non-normative* collective action that violates those norms [@wright_responding_1990; @becker_dynamic_2015]. Some instead refer to the two kinds of collective action as moderate and radical collective action [e.g., @jimenez-moya_by_2015]. Broadly, this distinction captures the intuition that there are some forms of collective action that are generally considered acceptable and others that are not.

Recent research has shown that this distinction matters for how people respond to collective action. @shuman_disrupting_2021 showed that non-normative, non-violent collective action is more effective than either normative or violent collective action at gaining concessions from those most resistant to social change. @feinberg_activists_2020 demonstrated, however, that observers are less supportive of social movements that use extreme protest actions. Similarly, @teixeira_is_2020 found that advantaged-group members perceive non-normative collective action by disadvantaged-group members as more damaging to their ingroup's social image and are thus less supportive of such action. @verkuyten_intolerance_2022 found that, in general, the public tends to be intolerant of non-normative collective action. This means that activists face the dilemma that non-normative collective action might be most effective at gaining concessions but that it also risks reducing popular support for a cause. Together, these studies show that the distinction between normative and non-normative collective action is psychologically and societally consequential.

These and other studies have, for the most part, relied on ad-hoc distinctions between what researchers themselves considered normative and non-normative collective action. In liberal democracies, researchers tend to consider actions such as signing petitions, voting in elections, or peaceful protest to be normative and actions such as blocking traffic, damaging property, or violent protest to be non-normative. But, as reactions to Black athletes kneeling in protest suggest, the same action might be considered acceptable by some and unacceptable by others. That is, the distinction between normative and non-normative collective action might be in the eye of the beholder.

In this research, we move this distinction from the researcher's intuition into the realm of scientific investigation and test for double standards in where people draw the line between acceptable (normative) and unacceptable (non-normative) forms of collective action. For that purpose, we define a double standard as judging the same action as more or less acceptable depending on who is performing the action and for what reason [for an analogous definition, see @foschi_double_2000].^[@feinberg_activists_2020 provided incidental evidence for double standards in judging collective action as, in a manipulation check, Black participants rated Black Lives Matter protests as less extreme than White participants and liberal participants rated protest actions for progressive causes as less extreme than conservative participants.] In this section, we derive predictions from social psychological theories about how identity-based and ideology-based double standards could lead people to judge the same protest actions as more or less acceptable depending on who the protesters are and what they are protesting.

@@

One domain in which ingroup bias manifests is judgments about other people's actions. @hewstone_ultimate_1990 reviewed research showing that people make ingroup-serving causal attributions when judging actions by ingroup and outgroup members, for example, by attributing negative behavior by outgroup members to internal causes but attributing negative behavior by ingroup members to external causes [see @pettigrew_ultimate_1979]. @valdesolo_moral_2007 demonstrated that participants judged the same selfish action as less unfair when performed by themselves or an ingroup member than when performed by an outgroup member. Other studies [@abrams_double_2013; @endevelt_everyone_2021; for exceptions, see @mendoza_for_2014; @pinto_membership_2010] provided further evidence for ingroup bias in judging moral transgressions by ingroup and outgroup members.

@@

System Justification Theory [@jost_decade_2004; for a recent review, see @jost_theory_2020] argues that people are motivated to defend, justify, and bolster the prevailing social, economic, and political system because doing so serves basic epistemic, existential, and relational needs. Both advantaged and disadvantaged groups are thought to be motivated to justify the existing system, although the strength of this motivation and its expression vary across individuals and situations.

@@

Just as system justification affects support for, and opposition to, collective action, we propose that it results in ideology-based double standards in what actions people consider acceptable forms of collective action. First, we hypothesize that  people will judge the same protest actions as more acceptable if the protest supports, rather than challenges, the system (*Hypothesis 2a*). This hypothesis reflects the assumption that the motivation to justify the system is, to some extent, universal. Second, we hypothesize that people will judge the same protest actions as more acceptable if they endorse system-justifying beliefs and the protest supports the system or if they reject system-justifying beliefs and the protest challenges the system (*Hypothesis 2b*). Notably, this hypothesis implies a symmetry in how people with different ideological orientations judge system-supporting and system-challenging collective action.

@@

Item response theory is a conceptual and statistical framework for understanding how the characteristics of both items and respondents shape responses to a set of items [@demars_irt_2010]. <!-- As such, it allows us to disentangle how acceptable protest actions are from how accepting respondents are of them and, in so doing, to detect double standards in judging collective action. --> When respondents judge whether each of a set of actions is an acceptable means of protest (1 = *yes*, 0 = *no*), a two-parameter logistic item response theory model estimates responses as a function of several latent (unobserved) parameters: 
$$ \Pr ( y_{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) $$
where $\Pr ( y_{ij} = 1 )$ is the probability that respondent $j$ considers action $i$ an acceptable means of protest, $\theta_j$ estimates how *accepting* respondent $j$ is of various protest actions, $\beta_i$ estimates how *acceptable* action $i$ is, and $\alpha_i$ estimates how *discriminating* action $i$ is.

@@

In this research, item response theory serves two purposes. First, item response theory provides a statistical framework for scale development. In Studies 1 and 2, we select the most discriminating and informative actions to build an instrument for capturing double standards in judging collective action. Second, item response theory provides a formal and statistical definition of double standards in judging collective action as respondents being more accepting of the same protest actions depending on who the protesters are and what they are protesting. In Experiments 1--3, we operationalize double standards as the differences in $\theta_{(z)}$ between conditions which, as $\theta_{(z)}$ is $z$-standardized, correspond to Cohen's $d$ effect sizes.

@@

The first purpose of our research is to develop an instrument to measure where people draw the line between acceptable and unacceptable forms of collective action. In Study 1, we ask participants to generate more and less extreme protest actions and compile a pool of protest actions for scale development. In Study 2, we ask participants to rate how acceptable they consider each of those protest actions to be and use item response theory to select the most discriminating and informative protest actions for our instrument. In so doing, we build an instrument that enables both present and future research to capture double standards in judging collective action. <!-- In this way, our research situates the distinction between normative and non-normative collective action in the eye of the beholders and opens it to scientific investigation. -->

@@

@@

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that, in line with *Hypothesis 1a* and *1b*, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members. Assuming that collective action by lower-status group members is seen as more threatening to the unequal status quo than collective action by higher-status group members and that people are, to same extent, motivated to justify the system, we tested the preregistered hypothesis that, in line with *Hypothesis 2a*, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group than if they were from a lower-status group. In non-preregistered analyses, we further explored whether participants would judge the same protest actions as more or less acceptable depending on their own ideological positions. In this way, Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

We recruited 515 participants from the Prolific subject pool who were UK citizens, 25 years old or older, and not current students.^[We had preregistered a sample of 500 participants, before exclusions, but included another 15 participants who completed the study without returning an approval code to the Prolific platform. Participants received, on average, \pounds 9.00 (\$10.93) per hour of participation.] As preregistered, we excluded 71 participants who failed an attention check. This resulted in a final sample of 443 participants ($\textit{Mdn} = 41$ years, age range: 25--76 years; 272 women, 171 men, 1 non-binary). Of these, 210 participants considered their past, current, or future jobs to be working-class jobs. Participants in this lower-status group did not have a university degree and placed themselves on the bottom three ranks of the subjective socioeconomic status ladder. Another 233 participants considered their past, current, or future jobs to be middle-class/professional jobs. Participants in this higher-status group had at least an undergraduate degree and placed themselves on the top four ranks of the subjective socioeconomic status ladder.

@@

We assessed reactions to the vignette by asking participants how outraged they would be if the government were to introduce this bill in real life, to what extent it would affect them personally, and to what extent it would affect people like them (1 = *not at all*, 7 = *very much*). We also asked participants to what extent they identified with people with the kinds of jobs most affected by the proposed bill (1 = *not at all*, 7 = *very much*) and how often, if at all, they had participated in protest actions such as the ones we asked about (1 = *never*, 5 = *very often*).

@@

Figure \ref{fig:f3} (B) the $z$-standardized propensity to consider more controversial protest actions acceptable as a function of the three ideological orientation variables. We found that participants who reported a more right-wing political orientation ($\beta_{xy} = -0.25, [-0.37, -0.15]$) and who expressed more agreement with system-justifying beliefs ($\beta_{xy} = -0.20, [-0.30, -0.10]$) tended to find fewer collective actions to be acceptable. In contrast, we found that, after controlling for the other two variables, social dominance orientation was not associated with participants' judgments about how acceptable various collective actions are ($\beta_{xy} = -0.05, [-0.15, 0.05]$). Overall, these findings suggest that people who are right-wing and endorse system-justifying beliefs tend to find various collective actions to be less acceptable than people who are left-wing and do not endorse system-justifying beliefs.^[We ran additional analyses to explore how these associations differed across experimental conditions. We found uncertainty intervals for these associations to overlap across conditions, suggesting that we did not have enough data to differentiate these varying effects.]

@@

Experiment 1 did not, however, provide a conclusive test of our hypotheses. First, by focusing on fictitious government bills, we might have chosen a scenario too far removed from current affairs to evoke strong ingroup bias in participants' judgments of protest actions. That said, participants' reactions to the experimental manipulation suggested that it evoked outrage and was understood in ingroup--outgroup terms. Second, by focusing on collective action to defend workers' rights, we arguably studied reactions to collective action for a progressive cause. To test for ideology-based double standards, we relied on the assumption that collective action by lower-status group members would be seen as more challenging to the prevailing system than collective action by higher-status group members. This assumption, however, was contradicted by our observation that participants who endorsed system-justifying beliefs tended to find actions by lower-status and higher-status protesters to be less acceptable. Experiment 2 addressed those limitations by focusing on a scenario we expected to evoke stronger responses from participants and by examining reactions to both system-challenging and system-defending collective action.

@@

In Experiment 2, we investigated potential double standards in judging collective action for or against defunding the police in the United States. That is, we tested whether Black and White Americans applied different standards when judging collective action taken by either Black or White protesters to protest either for or against police divestment as a possible solution to racist police violence. We conducted this study in January 2021, a time when, after months of unprecedented protests for racial justice, most Americans could be expected to be aware of, and have formed an opinion on, the Black Lives Matter movement. We focused on defunding the police as this position remained controversial among both Black and White Americans even as they broadly agreed on the need for police reform.^[In December 2020, only 34% of Black Americans and 18% of White Americans supported defunding the police while majorities of Black and White Americans supported police reforms such as eliminating qualified immunity (50% and 57%) or banning chokeholds [79% and 61%, @yougov_police_2020].]

@@

As expected, Black participants ($M = 6.18, \textit{SD} = 1.31$) reported being more outraged about recent incidents of police violence than White participants ($M = 5.56, \textit{SD} = 1.61$; Cohen's $d = 0.42$). Black participants strongly identified with their ingroup ($M = 6.46, \textit{SD} = 1.09$) and, across experimental conditions, identified more with Black protesters ($M = 4.66, \textit{SD} = 1.99$) than with White protesters ($M = 4.03, \textit{SD} = 2.16$; $d = 0.30$). In contrast, White participants identified less strongly with their ingroup than Black participants ($M = 4.85, \textit{SD} = 1.53$; $d = 1.21$) and did not identify more with White protesters ($M = 3.58, \textit{SD} = 2.01$) than with Black protesters ($M = 3.81, \textit{SD} = 1.77$; $d = -0.12$). Both White ($d = 0.67$) and Black ($d = 0.53$) participants tended to identify more with protesters protesting *for* defunding the police. On average, participants tended to describe their political orientation as moderately liberal ($M = 2.92, \textit{SD} = 1.73$) but, as expected, were divided in their support for defunding the police ($M = 3.37, \textit{SD} = 1.42$). That is, 56% supported defunding the police, 30% opposed defunding the police, and 14% remained undecided.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a function of their *z*-standardized endorsement of system-justifying beliefs. As preregistered, we modeled this relationship with two fixed effects, one estimating the effect of system-justifying beliefs on judgments about system-defending protest actions and one estimating their effect on judgments about system-challenging protest actions, and one varying effect estimating its variance across conditions. Supporting *Hypothesis 2b*, participants who *rejected* system-justifying beliefs were *more* likely to consider system-challenging protest actions (for defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "for"])`). We did not, however, find evidence for the ideological symmetry implied by *Hypothesis 2b* since participants who *endorsed* system-justifying beliefs were *not* more likely to consider system-defending protest actions (against defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "against"])`).

@@

We explored whether group identification moderated how the participants' and the protesters' group memberships affected participants' responses. To that end, we extended Model 1 by estimating participants' responses as a function of their *z*-standardized identification with their racial ingroup. As in Model 2, we modeled this relationship with a fixed and a varying effect. We found, however, that even participants who strongly identified with their racial ingroup (+1*SD*) did not consider protest actions performed by their ingroup to be more acceptable than the same actions performed by the relevant outgroup (`r with(m5, d_text[contrast == "h1a_contrast" & x_ide == 1L])`) and did not consider protest actions for a cause that was aligned with their ingroup's interests to be more acceptable than protest actions for a cause not aligned with their ingroup's interests (`r with(m5, d_text[contrast == "h1b_contrast" & x_ide == 1L])`). Our non-preregistered analyses thus suggested that group identification did not moderate group differences in judgments about collective action by ingroup and outgroup members.

@@

In Experiment 3, we investigated potential double standards in judging collective action for or against restricting abortion in the United States. That is, we tested whether liberal and conservative women and men applied different standards when judging collective action opposing or supporting further restricting abortion. We conducted this study in June 2022 when many expected the Supreme Court to overturn its *Roe v. Wade* decision that hitherto had guaranteed access to legal abortions in the first 20--24 weeks of pregnancy in the United States. We focused on restricting abortion as it is a conservative issue, as it primarily affects women's reproductive rights, and as the anticipated Supreme Court decision would prompt several states to restrict or ban abortion.

By comparing participants of different political orientations and by varying the protesters' cause, we tested the preregistered hypothesis that, in line with *symmetrical* ideology-based double standards, conservatives would judge the same protest actions to be more acceptable if the protesters supported restricting abortion and that liberals would judge the same protest actions to be more acceptable if the protesters opposed restricting abortion. We tested the alternative hypothesis that, in line with *asymmetrical* ideology-based double standards, conservatives would judge all protest actions to be equally unacceptable while liberals would judge the same protest actions to be more acceptable if the protesters opposed restricting abortion. By comparing male and female participants, we further tested the preregistered hypothesis that, in line with identity-based double standards, women would judge protest actions against restricting abortion to be more acceptable, on average, than men or protest actions for restricting abortion.

@@

We preregistered the sample size, hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822](https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822)). We deviated from the preregistration in two ways. First, we stopped collecting data before reaching the preregistered sample size. We did so as the Supreme Court decision on reproductive rights, announced on June 24, 2022, rendered the experimental manipulation invalid. Second, we do not report results for two further hypotheses---testing moral conviction as a potential mechanism underlying ideology-based double standards---as those findings are beyond the scope of this manuscript and will be reported in a future publication. We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)).

@@

We recruited 804 participants from the Prolific subject pool who were 18 years old or older and lived in the United States.^[Data were collected between June 10 and 23, 2022. Participants received, on average, \$24.70 per hour of participation.] We balanced the sample's gender and political composition so that half of all participants identified as conservative and the other half identified as liberal and so that roughly half of the participants in each group were women and men. As preregistered, we excluded 71 participants who failed at least one of five attention checks or reported a different political orientation than they had reported in the Prolific prescreening questionnaire. As noted above, we had to stop collecting data before reaching the preregistered sample size of 800 eligible participants. This resulted in a final sample of 733 participants ($\textit{Mdn} = 37$ years, age range: 18--93 years; 610 White, 55 Asian, 31 mixed, 20 Black, 16 other) of whom 181 were conservative men, 190 were conservative women, 185 were liberal men, and 177 were liberal women.

@@

After answering where they would place themselves along the political spectrum, in line with their response in the Prolific prescreening questionnaire, participants read the following paragraph:
